---
title: "Simulations exploring alpha and beta diversity"
author: "George Savva, (QIB)"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

This file will include short simulations to understand how differences in community compositions are reflected in alpha and beta diversity metrics.

In each simulation I'll create some data with given characteristics then look at how this is reflected in the diversity measures.  This could be extended into power calculations if needed.

## Simulation 1

* We'll assume that there are 100 possible species in my communities.  

* There are two different classes of environments from which we can sample communities, 

* In samples from the first environment each species is present with probability 0.4.

* In samples from the the second environment each species is present with probability 0.5.

* We'll sample ten communities per class, and look at their alpha and beta diversities

First sample a matrix of the community compositions:


```{r }

N = 10 # Number of samples per class
n = 100 # Number of possible species
set.seed("19122023")
classes <- rep(c("A", "B"), each=N)
classProbabilities <- c("A"=0.4,"B"=0.5)
dat <- sapply(classes, \(class) rbinom(n=n, size = 1, p=classProbabilities[class])) |> t()

# Check it looks OK
dat[1:10,1:10]
```

Now look at the alpha diversity (Shannon index):

```{r message=FALSE}

library(vegan)
library(ggplot2)
## Shannon index
t.test(diversity(dat) ~ classes)
ggplot() + aes(y=diversity(dat), x=classes) + 
  geom_boxplot()  +
  geom_point() + 
  stat_summary()

```

Beta-diversity (NMDS following Jaccard diversity):

```{r }

## Not so clear with the beta-diversity
vegdist(dat, method = "jaccard", binary = TRUE) |> 
  metaMDS(trace = 0) |>
  scores() |> 
  plot(col=factor(classes), pch=20)
  legend("topright", legend=c("A","B"), col=1:2,pch=20)

dm1 <- vegdist(dat, method = "jaccard", binary = TRUE)
adonis2(dm1 ~ classes  )  
  
anosim(dm1, grouping = classes)
```

So in this case we see a clear difference in alpha diversity between sites but it's not so clear for beta diversity.  It looks like the environment B samples are slightly more clustered while the A samples are more spread out.  The difference is not reflected in a PERMANOVA (via adonis2) or ANOSIM.

This is a good reminder that important differences in community composition might not be immediately visible in beta diversity plots.  Maybe you all know that anyway but it was a good reminder for me.

